[Artifact Presentation] Damegender: \\ Writing and Comparing Gender Detection Tools

### 2020-05-16 paper2sattose.tex
Reviewed up to section 4

\begin{abstract}
%Context
Diversity in software development teams have been identified as one of the main ingredients of a more productive, more healthy software community.
Thus, the interest of the research community in identifying who is contributing has increased in the last years.
In the software domain, and although other types of diversity exist, this is especially important for the case of gender.
% Objective, 
Given the large amount of publicly available data on the software development process that can be retrieved and analyzed from the Internet (e.g., GitHub, StackOverflow), the importance of having methods and tools that help with large amounts of data would be desirable.
% Method
In this paper we present a free software tool, called \texttt{damegender}, which we have conceived to given a name outputs the gender and a probability.
\texttt{damegender} is based on open databases from official census and uses Machine Learning to guess strings not classified as names, such as diminutives or nicknames.
% Results 
We have compared \texttt{damegender} with other tools, obtaining good results.
% Conclusions.
We hope the \texttt{damegender} tool can become a cornerstone for the scientific advancement of the study of gender (including gender gap) and diversity in the IT, and in particular in the free software community.
\end{abstract}

@@

In recent times, many research investigations have been made on gender diversity in the IT domain.

Examples of these efforts range from participation in Twitter~\cite{burger2011discriminating,mislove2011understanding},
in Wikipedia ~\cite{antin2011gender,hill2013wikipedia}, in science~\cite{holman2018gender,dollar1999gender}, and more specifically in the software domain in StackOverflow~\cite{vasilescu2012gender}, GitHub~\cite{vasilescu2015gender} and in Free/Libre/Open Source Software development~\cite{robles2014floss}.

The interest on gender diversity is become more and more relevant, and so does the identification methods that allow to perform comprehensive studies on gender representation in different domains, given the large amounts of data available, in particular from collaborative environments.

There are different ways to detect gender from a person name and perhaps a surname.
A first, more rudimentary, is based on data extracted from the census, Wikipedia, self-references in trust websites, searches in Google Images, among others.
So, for instance, in some studies, for example, about Twitter or GitHub, some people have used not only names to detect gender. 
Thus, we can find gender detection tools that infer the gender from faces in images~\cite{ranjan2017hyperface}, from hand-written annotations~\cite{liwicki2011automatic}, or from speeches~\cite{koppel2002automatically}.

Another way to do it is by using one of the existing Application Programming Interfaces (APIs).
This paper is about the latter, about their possibilities and limitations.
Therefore, (i) we evaluate the quality and price of different commercial solutions;
(ii) we discuss about solutions using free licenses;
(iii) we investigate what happens with those names without census, for example, nicknames or diminutives; and
(iv) we elaborate on how massive gender detection from Internet, for example, mailing
  lists or software repositories, can be done.

@@

(ii) a tool, called \texttt{damegender}, guessing gender from a name
  giving support to Spanish and English from the open data census
  provides by the states built to understand current technologies in
  detail; this tools has been compared with APIs using an international
  dataset with good results.

@@

(iv) a proof-of-concept of \texttt{damegender} to detect
  gender in mailing lists and software repositories.

@@

In Section 2 we explain \texttt{damegender} as solution to the
problem of name-to-gender inference. Section 3 introduces the Open Datasets used as the \emph{golden truth}.
Section 4 presents a feature comparison with other name-to-gender inference services and tools.
Section 5 reports on values of accuracy and offers a confusion matrix using a scientific dataset.
Section 6 is about how we use Machine Learning in \texttt{damegender}.
Section 7 discusses limitations and further research, and concludes the paper.

\section{Damegender}

\texttt{damegender}\footnote{https://github.com/davidam/damegender} is a gender detection tool under a Free Software license (in particular, the GNU General Public License v3.0). 
It has been implemented in Python to take advantage of many other free software tools used in the scientific domain, such as the Natural Language Toolkit (NLTK) for Natural Language Processing~\cite{loper2002nltk}, Scikit for Machine Learning~\cite{pedregosa2011scikit}, Numpy for Numerical Computation~\cite{van2011numpy}, and Matplotlib to visualize results~\cite{hunter2007matplotlib}. 
At its current point it is linked to Perceval~\cite{duenas2018perceval}, a tool specialized in retrieving and gathering data from software repositories, such as git and mailing lists.
\texttt{damegender} is a Python package that can be installed using PIP (the package installer for Python) from the console.

@@

The main reason for developing \texttt{damegender} is that there are not many free software tools that help in the identification of gender. Before \texttt{damegender}, only \texttt{Gender guesser}\footnote{https://github.com/lead-ratings/gender-guesser} offered a free software solution in this field~\cite{krawetz2006gender}, and the project has not been active for more than three years now.
The best contribution of \texttt{Gender guesser} is the dataset containing 48,528 names with a good classification by countries\footnote{\url{https://raw.githubusercontent.com/lead-ratings/gender-guesser/master/gender_guesser/data/nam_dict.txt}}.

\section{Datasets}

@@

There are several sources to create a databases, being the most common:  
(1) a census published with a free license (open census way), 
(2) a dataset released with a free license in a free software package (free software way), 
(3) a dataset retrieved from commercial APIs (commercial API way), and
(4) a dataset which is the result of an investigation and that has been released publicly (scientific way).

In \texttt{damegender}, we are including Open Data census about names and gender, from institutions such as INE.es (the Spanish National Statistics Institute), or the governments of Uruguay, USA and United Kingdom. The datasets provided by the software package is incrementing the speed retrieving data.

@@

Hence, we are using the census approach as base of truth to distinguish if a name is male or female in a geographical area. 
Generally, a name has a strong weight to determine if it is a male or a female on this way.
For instance, David is registered 365,196 times as male, but 0 times as female in the data offered by the Spain National Institute of Statistics.
There are names that heavily depend on the region. 
For instance, Andrea would be considered a female name in Germany, but a male name in Italy.
However, many countries do not provide Open Data census about gender and names.

@@

We have evaluated to include data from the second option (datasets released with a free license). 
For instance, Natural Language Tool Kit offers 8,000 labeled English names classified as male or female. 
Another example is \texttt{Gender Guesser} a good dataset for international names with different categories to define the probability. 
The problem with these data is that we have observed that they do not have the quality of National Statistics Institutes.

@@

The third approach is based on the trust on commercial solutions, in the same way we trust search engines when we make searches in Internet.
This is because commercial APIs can be seen just a black box, so we do not know where the data comes from and how it has been treated.
As at this point, commercial APIs offer better results as other solutions, \texttt{damegender} gives the possibility to include data from them.
Thus, it is possible to download JSON files from the main commercial name-to-gender inference API solutions (e.g., \texttt{genderapi}, \texttt{genderize}, \texttt{namsor}, \texttt{nameapi}) and use it as the dataset.
There are certain uses that are currently only available in such tools.

@@

As a final goal, we envision to build a free dataset with names and gender, that builds on top of \texttt{Gender Guesser} and that can be made available as Wikidata. 
Perhaps, to complete this work, we need to combine an automated with a manual process as described in~\cite{10.7717/peerj-cs.156}.

\section{Feature comparison with other tools}

@@

Santamaria and Mihaljevic~\cite{10.7717/peerj-cs.156} offer a framework to classify gender tools.
The features observed in this framework are: (i) database size (as of January 2018), (ii) if there are regular data updates, (iii) if they handle unstructured full name strings, (iv) if they handle surnames, (v) if they handle non-Latin alphabets, (vi) if implicit geolocalization is available, (vii) if locale exists, (viii) the type of assignment, (ix) if free parameters are possible, (x) if they offer prediction, (xi) if the tool is released under an open source license, (xii) if they offer an API, (xiii) the amount of monthly free requests, and (xiv) the monthly subscription cost (calculated for 100,000 requests/month)).

@@

\section{Reproducing values of accuracy and confusion matrix}

@@

\section{Machine Learning}

@@

\begin{table}
\footnotesize
\center
\begin{tabular}[]{lcccc}
  \hline
  ML Algorithm & Acc & Prec & F1 & Recall \tabularnewline
  \hline
  %\endhead
 Support Vector Machines             &    0.879 &     0.972 &   0.972 &    1.0  \tabularnewline
 Random Forest                       &    0.862 &     0.902 &   0.902 &    1.0  \tabularnewline
 NLTK (Bayes)                        &    0.862 &     0.902 &   0.902 &    1.0  \tabularnewline
 Multinomial Navie Bayes             &    0.782 &     0.791 &   0.791 &    1.0  \tabularnewline
 Tree                                &    0.764 &     0.821 &   0.796 &    1.0  \tabularnewline
 Stoch. Gradient Distrib.    &    0.709 &     0.943 &   0.815 &    1.0  \tabularnewline
 Gaussian Naive Bayes                &    0.709 &     0.968 &   0.887 &    1.0  \tabularnewline
 Bernoulli Naive Bayes               &    0.699 &     0.965 &   0.816 &    1.0  \tabularnewline

@@

\section{Limitations and further research}

>>>>

\begin{abstract}
%Context
Diversity in software development teams has been identified as one of the main ingredients of a more productive, more healthy software community.
Thus, the interest of the research community in identifying who is contributing to a project has increased in the last years.
In the software domain, and although other types of diversity exist, this is particularly important for the case of gender.
% Objective, 
Given the large amount of publicly available data on the software development process that can be retrieved and analyzed from the Internet (e.g., GitHub, StackOverflow), the importance of having methods and tools that help to identify the participation of women within large amounts of data is desirable.
% Method name-to-gender inference
In this paper we present a free software tool, called \texttt{damegender}, which we have conceived to infer the gender from the name of a participant.
\texttt{damegender} is based on open databases from official census and uses Machine Learning.
An experimental branch is currently under development for those cases where instead of names we have diminutives or nicknames.
% Results 
We have compared \texttt{damegender} with other name-to-gender inference tools, obtaining good results.
% Conclusions.
We hope the \texttt{damegender} tool can become a cornerstone for the scientific advancement of the study of gender (including the gender gap) and diversity in the IT, and in particular in the free software community.
\end{abstract}

@@

In recent times, many research investigations have been made on gender diversity in the IT domain.
Examples of these efforts range from the participation in Twitter~\cite{burger2011discriminating,mislove2011understanding},
in Wikipedia~\cite{antin2011gender,hill2013wikipedia}, in science~\cite{holman2018gender,dollar1999gender}, and more specifically in the software domain in StackOverflow~\cite{vasilescu2012gender,vasilescu2015gender}, GitHub~\cite{vasilescu2015gender}, and free software development~\cite{robles2014floss,izquierdo2018openstack,lee2019floss,catolino2019gender}.

The interest on gender diversity has become more and more relevant, and so does the identification methods that allow to perform comprehensive studies on gender representation, given the large amounts of data available, in particular from collaborative environments.

The most used method to infer the gender of participants is by looking at their name.
However, there are some studies about Twitter or GitHub were not only names to detect gender. 
In them we can find gender detection tools that infer the gender from faces in images~\cite{ranjan2017hyperface}, from hand-written annotations~\cite{liwicki2011automatic}, or from speeches~\cite{koppel2002automatically}.

Currently, there are different ways to detect gender from a person's name (and maybe other personal data, such as the surname, the geographic location, etc.).
A first one is based on data extracted from the Wikipedia, self-references in trust websites, searches in Google Images, among others.
Another way to do it is by using one of the existing Application Programming Interfaces (APIs).
This paper is about the latter, focusing on their possibilities and limitations.
Therefore, (i) we evaluate the quality (and accessibility, including price) of different existing solutions;
(ii) we discuss about their limitations; and
(iii) we investigate what happens with those names not included in a census, for example, nicknames or diminutives.
% and (iv) we elaborate on how massive gender detection from Internet, for example, mailing lists or software repositories, can be done.

@@

(ii) a tool, called \texttt{damegender}, guessing gender from a name
  giving support to Spanish and English from the open data census
  provides by the states built to understand current technologies in
  detail; this tools has been compared with APIs using an international
  dataset with good results; and

@@

%(iv) a proof-of-concept of \texttt{damegender} to detect gender in mailing lists and software repositories.

@@

In Section~\ref{sec:damegender} we explain \texttt{damegender} as solution to the problem of name-to-gender inference.
Section~\ref{sec:datasets} introduces the Open Datasets used as the \emph{golden truth}.
Section~\ref{sec:comparison} presents a feature comparison with other name-to-gender inference services and tools.
Section~\ref{sec:reproducing} reports on values of accuracy and offers a confusion matrix using a scientific dataset.
Section~\ref{sec:ml} is about how we use Machine Learning in \texttt{damegender}.
Section~\ref{sec:limitations} discusses limitations and further research, and concludes the paper.

\section{Damegender}
\label{sec:damegender}

\texttt{damegender}\footnote{https://github.com/davidam/damegender} is a gender detection tool released under a free software license (in particular, the GNU General Public License v3.0). 
It has been implemented in Python to take advantage of many other free software tools used in the scientific domain, such as the Natural Language Toolkit (NLTK) for natural language processing~\cite{loper2002nltk}, Scikit for machine learning~\cite{pedregosa2011scikit}, Numpy nor mumerical computation~\cite{van2011numpy}, and Matplotlib to visualize results~\cite{hunter2007matplotlib}. 
At its current point it is linked to Perceval~\cite{duenas2018perceval}, a tool specialized in retrieving and gathering data from software repositories, such as git and mailing lists.
\texttt{damegender} is a Python package that can be easily installed using PIP (the package installer for Python) from the console.

@@

The main reason for developing \texttt{damegender} is that there are not many free software tools that help in the identification of gender. 
Before \texttt{damegender}, only \texttt{Gender guesser}\footnote{https://github.com/lead-ratings/gender-guesser} offered a free software solution in this field~\cite{krawetz2006gender}, and the project (i) has not been active for more than three years now, (ii) lies technologically well behind other solutions.
The best contribution of \texttt{Gender guesser} is the dataset containing 48,528 names with a good classification by countries\footnote{\url{https://raw.githubusercontent.com/lead-ratings/gender-guesser/master/gender_guesser/data/nam_dict.txt}}.

\section{Datasets}
\label{sec:datasets}

@@

There are several sources to create these databases, being the most common:  
(1) a census published with a free license (open census way), 
(2) a dataset released with a free license in a free software package (free software way), 
(3) a dataset retrieved from commercial APIs (commercial API way), and
(4) a dataset which is the result of an investigation and that has been released publicly (scientific way).

In \texttt{damegender}, we are including Open Data census about names and gender, usually offered by governments, such as the ones from Spain (by means of INE.es, the Spanish National Statistics Institute), Uruguay, USA and United Kingdom. 
%The datasets provided by the software package is incrementing the speed retrieving data.

@@

Generally, a name has a strong weight to determine if it is a male or a female using this procedure.
For instance, David is registered 365,196 times as male, but 0 times as female in the data offered by the Spain National Institute of Statistics.
There are names that heavily depend on the region. 
For instance, Andrea would be considered a female name in Germany, but a male name in Italy.
Hence, we are using the census approach as base of truth to distinguish if a name is male or female in a geographical area. 
However, many countries do not provide Open Data census about gender and names, although our hope is that they will do in the near future.

@@

We have evaluated to include data from the second option (datasets released with a free license). 
For instance, the Natural Language Tool Kit offers 8,000 labeled English names classified as male or female. 
Another example is \texttt{Gender Guesser}, which provides a good dataset for international names with different categories to define the probability. 
Although these datasets can be incorporated into \texttt{damegender}, the problem with them is that, in general, we have observed that they do not have the quality of National Statistics Institutes.

@@

The third approach is to create the dataset from existing, usually commercial solutions.
Here we have to trust their results, in the same way we trust search engines when we make searches in Internet.
This is because commercial APIs can be seen as a black box -- we do not know where the data comes from and how it has been treated.
But, at least, we can measure its quality, as we will do next.
So, we will see that --at this point-- commercial APIs offer better results than other solutions.
That is why \texttt{damegender} gives the possibility to include data from them.
It is possible to download JSON files from the main commercial name-to-gender inference API solutions (e.g., \texttt{genderapi}, \texttt{genderize}, \texttt{namsor}, \texttt{nameapi}) and use it as the dataset.
There are certain uses that are currently only available in such tools.

@@

As a final goal, we envision to build a free dataset with names and gender, that builds on top of \texttt{Gender Guesser} and that can be made available as Wikidata with a free license. 
Perhaps, to complete this work, we need to combine an automated with a manual process as described in~\cite{10.7717/peerj-cs.156}.

\section{Feature comparison with other tools}
\label{sec:comparison}

@@

Santamaría and Mihaljevic~\cite{10.7717/peerj-cs.156} offer a framework to classify gender tools.
The features observed in this framework are: (i) database size (as of January 2018), (ii) if there are regular data updates, (iii) if they handle unstructured full name strings, (iv) if they handle surnames, (v) if they handle non-Latin alphabets, (vi) if implicit geolocalization is available, (vii) if locale exists, (viii) the type of assignment, (ix) if free parameters are possible, (x) if they offer prediction, (xi) if the tool is released under an open source license, (xii) if they offer an API, (xiii) the amount of monthly free requests, and (xiv) the monthly subscription cost (calculated for 100,000 requests/month)).

@@

\section{Reproducing values of accuracy and confusion matrix}
\label{sec:reproducing}

@@

\section{Machine Learning}
\label{sec:ml}

@@

\begin{table}
\footnotesize
\center
\begin{tabular}[]{lcccc}
  \hline
  ML Algorithm & Acc & Prec & F1 & Recall \tabularnewline
  \hline
  %\endhead
 Support Vector Machines             &    0.879 &     0.972 &   0.972 &    1.0  \tabularnewline
 Random Forest                       &    0.862 &     0.902 &   0.902 &    1.0  \tabularnewline
 NLTK (Bayes)                        &    0.862 &     0.902 &   0.902 &    1.0  \tabularnewline
 Multinomial Naive Bayes             &    0.782 &     0.791 &   0.791 &    1.0  \tabularnewline
 Tree                                &    0.764 &     0.821 &   0.796 &    1.0  \tabularnewline
 Stoch. Gradient Distrib.    &    0.709 &     0.943 &   0.815 &    1.0  \tabularnewline
 Gaussian Naive Bayes                &    0.709 &     0.968 &   0.887 &    1.0  \tabularnewline
 Bernoulli Naive Bayes               &    0.699 &     0.965 &   0.816 &    1.0  \tabularnewline

@@

\section{Limitations and further research}
\label{sec:limitations}
